The role of advection versus diffusion in intraventricular transport

### 2025-03-18 paper/main.tex
add references intro

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. CSF is a clear liquid mostly produced by the choroid
plexus that flows through the ventricular system and subarachnoid
space~\cite{Damkier2013}. By transporting growth factors, nutrients, neurotransmitters
and metabolites as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{CITATTION}. Disruption of the
CSF circulation is detrimental to brain health and disturbs cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment, but for which there is 
little agreement on optimal treatments~\cite{flannery2014pediatric}. 
Of particular importance is the fact that
hydrocephalus is observed in animal models and humans with deficient motile
cilia~\cite{Afzelius2004Cilia, Olstad2019CiliaryDevelopment, Mitchison2017Motile},
indicating that the cilia play a crucial role for healthy brain functioning.

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia lining the ventricular walls, CSF secretion by the choroid
plexus, and pulsatile pressure gradients generated by the cardiac
cycle, respiration, neural activity, and bodily
movement~\cite{Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment}.
Motile cilia are evolutionarily conserved, microscopic hair-like
structures that beat in a metachronal wave pattern at frequencies of
10--30 Hz to mediate fluid flow~\cite{mitchell2007evolution, Ringers2020Role}.
Given their microscopic size, ranging from
6 to 15 microns, the cilia clearly contributes primarily to flow near the
ventricular surface in the human brain~\cite{Siyahhan2014FlowVentricles}.
On the other hand, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled}.

@@

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
First, we computed flow patterns and simulated solute transport with a computational model. To
this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies. Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages,
while still being well-conserved with mammals. 
Next, we validated our computation model with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as the ventricular geometry, greatly
impact solute distribution. Altogether, our work presents a computational
framework together with new concepts of how cilia contribute to distribution of molecules and
particles within the brain ventricles.

>>>>

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. CSF is a clear liquid mostly produced by the choroid
plexus that flows through the ventricular system and subarachnoid
space~\cite{Damkier2013}. By transporting growth factors, nutrients, neurotransmitters
and metabolites as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{Ringers2020Role, fame2020emergence, del2010ependymal}.
Disruption of the CSF circulation is detrimental to brain health and disturbs
cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment, but for which there is 
little agreement on optimal treatments~\cite{flannery2014pediatric}. 
Of particular importance is the fact that
hydrocephalus is observed in animal models and humans with deficient motile
cilia~\cite{Afzelius2004Cilia, Olstad2019CiliaryDevelopment, Mitchison2017Motile},
indicating that the cilia play a crucial role for healthy brain functioning.

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movement~\cite{del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Motile cilia are evolutionarily conserved, microscopic hair-like
structures that beat in a coordinate pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution, Ringers2020Role}.
Given their microscopic size, ranging from
6 to 15 microns, the cilia clearly contributes primarily to flow near the
ventricular surface in the human brain~\cite{Siyahhan2014FlowVentricles}.
On the other hand, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled}.

@@

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
First, we computed flow patterns and simulated solute transport with a computational model. To
this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies. Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
Next, we validated our computation model with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as the ventricular geometry, largely
impact solute distribution. Altogether, our work presents a computational
framework together with new concepts of how cilia distribute molecules and
particles within the brain ventricles.
